Linear Ordered Attribute Grammar scheduling using SAT-solving

### 2014-10-17 paper/paper.tex
Some comments by Atze

\begin{abstract}
Many computations over trees can be expressed using attribute
grammars. Compilers for attribute grammars need to find an evaluation
order (or schedule) in order to generate efficient code. For the class
of linear ordered attribute grammars such a schedule can be found
statically, but this problem is known to be NP-hard.

In this paper, we show how to encode linear ordered attribute grammar
scheduling as a SAT-problem. Cycle-freeness of the dependency graphs
is expressed by making the graphs chordal, which is a new approach for
encoding such graph properties in SAT.

@@

\section{Introduction}
Attribute Grammars \cite{knuth68} are a formalism for describing the semantics of context-free languages, thereby making them suitable for the construction of compilers. Examples of compilers written using attribute grammars are JastAddJ \cite{Ekman:2007}, a Java compiler, and UHC \cite{Dijkstra:2009}, a Haskell compiler. The UHC is also the motivation for this paper and is used as a real world test case for evaluating the effectiveness of the given approach.

An attribute grammar essentially describes computations over trees, also known as folds or catamorphisms. Although the attribute grammar defines exactly \emph{what} should be computed, it does not define \emph{when} values should be computed. It is therefore up to the attribute grammar compiler (the compiler that translates the attribute grammar definition into a runtime evaluator) to find an evaluation order. Such an evaluation order or schedule should satisfy the dependencies from the attribute grammar definition and must be found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps. Specifically, no cyclic data dependencies may occur at runtime for any given parse tree.

@@

\subsection{Summary}
We solve the linear ordered attribute grammar scheduling problem by translating it to the Boolean satisfiability problem (SAT), one of the best-known NP-complete problems \cite{Cook:1971}. The SAT problem has been the subject of a lot of research and even though the worst case runtime of all known SAT-solving algorithms is exponential in the input size, many SAT-solvers work very well in practice. \Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating to SAT we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community. In our implementation we use MiniSat\footnote{\url{http://minisat.se}} \cite{Een:2004}.

@@

Apart from the fact that this translation to SAT helps in efficiently (in practice) solving the scheduling problem, there also is another benefit: it is now possible to encode extra constraints on the resulting schedule. We show two scheduling optimisations that are interesting from an attribute grammar perspective, for which the optimal schedule can be found efficiently by expressing the optimisation in the SAT problem.

@@

The outline of the paper is as follows. We start in \sectref{sect:ag} and \sectref{sect:loag} by explaining linear ordered attribute grammars and the difficulties in scheduling in more detail. We then describe the translation to SAT in \sectref{sect:translation} and show the effectiveness of our approach in \sectref{sect:results}. In \sectref{sect:optimisations} we describe the two attribute grammar optimisations and finally in \sectref{sect:conclusion} we discuss some problems and conclude.

@@

For dynamic scheduling one could rely on existing machinery for lazy evaluation, for example in Haskell, which is the approach taken by \cite{saraiva99}. There the attribute grammar definitions are translated to lazy Haskell functions in a straightforward way by producing functions that take the inherited attributes as argument and return the synthesized attributes. Whenever an inherited attribute (indirectly) depends on the value of a synthesized attribute, this means the function has a cyclic definition, which is no problem for languages with lazy semantics, and can actually result in efficient code \cite{Bird:1984}.

@@

\section{Translation to SAT} \label{sect:translation}
To represent the scheduling problem as a Boolean formula we introduce a variable for each edge, indicating the direction of the edge. The direct dependencies coming from the source code are constants, but for the rest of the edges the SAT solver can decide on the direction. However, the encoding has been chosen in such way that a valid assignment of the variables corresponds to a valid schedule.

@@

In the translation to the SAT-problem we have used a novel technique for ruling out cycles in the SAT problem using chordal graphs. This technique makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions. We believe that this technique is applicable in other problems using SAT-solvers as well.

>>>>

\begin{abstract}
Many computations over trees can be specified using attribute
grammars. Compilers for attribute grammars need to find an evaluation
order (or schedule) in order to generate efficient code. For the class
of linear ordered attribute grammars such a schedule can be found
statically, but this problem is known to be NP-hard.

In this paper, we show how to encode linear ordered attribute grammar
scheduling as a SAT-problem. For such grammars it is necessary to ensure that the dependency graph is cycle free, which we approach in a novel way by transforming the dependency graph to a chordal graph allowing the cycle freeness to be efficiently expressed and computed using SAT solvers.

@@

\section{Introduction}
Attribute Grammars \cite{knuth68} are a formalism for describing the semantics of context-free languages, thereby making them suitable for the construction of compilers. Examples of compilers written using attribute grammars are JastAddJ \cite{Ekman:2007}, a Java compiler, and UHC \cite{Dijkstra:2009}, a Haskell compiler. The use of attribute grammars in the UHC also motivates this paper and as such UHC is a real world test case for evaluating the effectiveness of the given approach.

An attribute grammar essentially describes computations over trees, also known as folds or catamorphisms. Although the attribute grammar defines exactly \emph{what} should be computed, it does not define \emph{when} values should be computed. It is therefore up to the attribute grammar compiler (the compiler that translates the attribute grammar definition into a runtime evaluator) to find an evaluation order. Such an evaluation order or schedule should satisfy the dependencies induced by the specification of attribute computations in the attribute grammar definition and must be found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps. Specifically, no cyclic data dependencies may occur at runtime for any given parse tree.

@@

\subsection{Summary}
We solve the linear ordered attribute grammar scheduling problem by translating it into the Boolean satisfiability problem (SAT), one of the best-known NP-complete problems \cite{Cook:1971}. The SAT problem has been the subject of a lot of research and even though the worst case runtime of all known SAT-solving algorithms is exponential in the input size, many SAT-solvers work very well in practice. \Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating into SAT we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community. In our implementation we use MiniSat\footnote{\url{http://minisat.se}} \cite{Een:2004}.

@@

Apart from the fact that this translation into SAT helps in efficiently (in practice) solving the scheduling problem, there also is another benefit: it is now possible to encode extra constraints on the resulting schedule. We show two scheduling optimisations that are interesting from an attribute grammar perspective, for which the optimal schedule can be found efficiently by expressing the optimisation in the SAT problem.

@@

The outline of the paper is as follows. We start in \sectref{sect:ag} and \sectref{sect:loag} by explaining linear ordered attribute grammars and the difficulties in scheduling in more detail. We then describe the translation into SAT in \sectref{sect:translation} and show the effectiveness of our approach in \sectref{sect:results}. In \sectref{sect:optimisations} we describe the two attribute grammar optimisations and finally in \sectref{sect:conclusion} we discuss some problems and conclude.

@@

For dynamic scheduling one could rely on existing machinery for lazy evaluation, for example in Haskell, which is the approach taken by \cite{saraiva99}. There the attribute grammar definitions are translated into lazy Haskell functions in a straightforward way by producing functions that take the inherited attributes as argument and return the synthesized attributes. Whenever an inherited attribute (indirectly) depends on the value of a synthesized attribute, this means the function has a cyclic definition, which is no problem for languages with lazy semantics, and can actually result in efficient code \cite{Bird:1984}.

@@

\section{Translation into SAT} \label{sect:translation}
To represent the scheduling problem as a Boolean formula we introduce a variable for each edge, indicating the direction of the edge. The direct dependencies coming from the source code are constants, but for the rest of the edges the SAT solver can decide on the direction. However, the encoding has been chosen in such way that a valid assignment of the variables corresponds to a valid schedule.

@@

In the translation into the SAT-problem we have used a novel technique for ruling out cycles in the SAT problem using chordal graphs. This technique makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions. We believe that this technique is applicable in other problems using SAT-solvers as well.
